---
title: "ARDL_model"
author: "BOUILLON Mathis"
date: "2025-01-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Loading packages : 
```{r}
# Load necessary packages for time series analysis
if (!require(tseries)) install.packages("tseries")
library(tseries)
```


Loading the final database : 

```{r}
final_database <- read_csv("final_database.csv", show_col_types = FALSE)
final_database$Date <- as.Date(final_database$Date)

numeric_columns <- setdiff(names(final_database), "Date")  # Sélectionner toutes les colonnes sauf 'Date'
final_database[numeric_columns] <- lapply(final_database[numeric_columns], as.numeric)

```

# ARDL model 

First, we want to check the hypothesis of this model. 

THen, we examine the stationarity of each variable using a KPSS test. 
```{r}
# Keep only numeric columns for stationarity testing
df_numeric <- final_database[, sapply(final_database, is.numeric)]

# Prepare a data frame to store stationarity test results
stationarity_results_kpss <- data.frame(
  Variable = colnames(df_numeric),
  p_value_Level = NA,  # p-value at the level
  p_value_Diff = NA,   # p-value after differencing
  Stationarity = NA    # Final stationarity classification
)

# Perform KPSS test on each variable for stationarity
for (i in 1:ncol(df_numeric)) {
  var_name <- colnames(df_numeric)[i]
  variable <- df_numeric[[var_name]]
  
  # KPSS test at the level
  kpss_level <- kpss.test(na.omit(variable), null = "Level")
  p_value_level <- kpss_level$p.value
  
  # KPSS test after differencing
  diff_variable <- diff(variable, differences = 1)
  kpss_diff <- kpss.test(na.omit(diff_variable), null = "Level")
  p_value_diff <- kpss_diff$p.value
  
  # Determine the stationarity level
  if (p_value_level > 0.05) {
    status <- "I(0)"
  } else if (p_value_diff > 0.05) {
    status <- "I(1)"
  } else {
    status <- "I(2)"
  }
  
  # Save results in the data frame
  stationarity_results_kpss[i, ] <- c(var_name, p_value_level, p_value_diff, status)
}

# Display KPSS stationarity test results
print(stationarity_results_kpss)

```

We only select variables being I(0) or I(1)

```{r}
# Select variables that are either I(0) or I(1)
i0_i1_vars <- stationarity_results_kpss$Variable[
  stationarity_results_kpss$Stationarity %in% c("I(0)", "I(1)")
]

# Ensure variable names match those in the original database
i0_i1_vars <- intersect(colnames(final_database), i0_i1_vars)

# Filter the data to include only Date and selected variables
filtered_df <- final_database[, c("Date", i0_i1_vars), drop = FALSE]
```


Then, we perform the Granger causality test : 

```{r}
# Load library for Granger causality tests
library(lmtest)

# Identify explanatory variables (exclude "Date" and "Return")
variables <- colnames(filtered_df)[!(colnames(filtered_df) %in% c("Date", "Return"))]

# Initialize a list to store Granger causality results
bilateral_results <- list()

# Test Granger causality in both directions for each variable
for (var in variables) {
  cat("\n--- Granger test between Return and", var, "---\n")
  
  # Check if the variable Granger-causes the target
  test1 <- grangertest(filtered_df$Return ~ filtered_df[[var]], order = 1)
  
  # Check if the target Granger-causes the variable
  test2 <- grangertest(filtered_df[[var]] ~ filtered_df$Return, order = 1)
  
  # Save results in the list
  bilateral_results[[var]] <- list(
    "Variable -> Target" = test1,
    "Target -> Variable" = test2
  )
  
  # Display results for each direction
  cat("\n---", var, "Granger-causes Return ---\n")
  print(test1)
  cat("\n--- Return Granger-causes", var, "---\n")
  print(test2)
}

```


The only variable that is not Granger-caused by the variable Return is the inflation; then, we can only use that one in our ARDL model 

```{r}
# Load ARDL package and subset data for model building
library(ARDL)
new_dataframe <- final_database[, c("Date", "Return", "CPIH_YTYPCT")]

# Split data into training and testing sets
new_dataframe_train <- new_dataframe[new_dataframe$Date < "2023-01-01", ]

# Fit ARDL model and identify the best model using AIC
models_new <- auto_ardl(Return ~ CPIH_YTYPCT, data = new_dataframe_train, max_order = 10)

# Display the top 20 models based on AIC
models_new$top_orders

```

The best model based on AIC is with 3 lags for both variables. 
Then, we run this model : 

```{r}
# Extract and summarize the best ARDL model
ardl_new <- models_new$best_model
ardl_new$order
summary(ardl_new)

# Retrieve and visualize residuals of the model
residuals_ardl_new <- residuals(ardl_new)

# Plot residuals over time
par(mfrow = c(2, 2))  # Create a 2x2 grid for plots
plot(residuals_ardl_new, type = "l", main = "Residuals", ylab = "Residuals", xlab = "Observations")
abline(h = 0, col = "red")  # Add a horizontal line at zero

# Plot histogram of residuals
hist(residuals_ardl_new, main = "Residuals Histogram", xlab = "Residuals", col = "blue", border = "black", breaks = 50)

# Perform Durbin-Watson test for autocorrelation in residuals
library(lmtest)
dwtest(ardl_new)

# Plot autocorrelation of residuals
acf(residuals_ardl_new, main = "Residuals Autocorrelation")

```
We observe that not so many coefficients are significant; it was predictable given that we did not put enough variables in  our model. The one left (inflation) is too de-correlated with the target to have a significant impact. 
However, the residuals are not correlated (DW close to 2), and the distribution of the residuals looks like a normal one. We can still observed a high volatility. 

Now, we want to check if the variables are co-integrated. As their order is different, we use the cointegration Bound Test after running the ARDL model : 

```{r}
# Perform bounds F-test for cointegration
bounds_f_test(ardl_new, case = 2)
bounds_f_test(ardl_new, case = 3)
```

There is a possible cointegration given the p-value. 

Then, we can perform an ECM : 

```{r}
# Fit an Unrestricted Error Correction Model (UECM)
uecm_new <- uecm(ardl_new)
summary(uecm_new)

# Fit a Restricted Error Correction Model (RECM)
recm_new <- recm(uecm_new, case = 2)
summary(recm_new)

# Display short-run and long-run multipliers
multipliers(ardl_new, type = "sr")  # Short-run multipliers
multipliers(ardl_new)               # Long-run multipliers

```


To perform prediction, we transform our model into a linear model 
```{r}
# Convert ARDL model to linear model for forecasting
ardl_new_lm <- to_lm(ardl_new)

# Predict using in-sample data
insample_data <- ardl_new$model
predicted_values <- predict(ardl_new_lm, newdata = insample_data)

# Add predictions to the training data frame
used_rows <- rownames(insample_data)
new_dataframe_train$Predicted <- NA
new_dataframe_train$Predicted[as.numeric(used_rows)] <- predicted_values
```


We plot predictions on graph to compare it to the observed returns of CAC 40 index : 

```{r}
# Plot observed vs predicted returns by 5-year periods
library(dplyr)
library(ggplot2)

# Add a column for 5-year periods
plot_data_train <- new_dataframe_train %>%
  mutate(
    Year = as.numeric(format(Date, "%Y")),
    Period = paste0(floor(Year / 5) * 5, "-", floor(Year / 5) * 5 + 4)
  )

# Generate plots for each 5-year period
periods <- unique(plot_data_train$Period)
for (period in periods) {
  period_data <- filter(plot_data_train, Period == period)
  
  p <- ggplot(period_data, aes(x = Date)) +
    geom_line(aes(y = Return), color = 'black', linewidth = 1) +
    geom_line(aes(y = Predicted), color = 'blue', linewidth = 1) +
    labs(title = paste("ARDL Predictions - Period:", period), x = "Date", y = "Return") +
    theme_minimal() +
    theme(legend.position = "none")
  
  print(p)
}

# Compute performance metrics (RMSE, MAE) for the training set
errors_train <- plot_data_train$Return - plot_data_train$Predicted
rmse_train <- sqrt(mean(errors_train^2, na.rm = TRUE))
mae_train <- mean(abs(errors_train), na.rm = TRUE)

cat("Training set performance:\n")
cat("RMSE: ", rmse_train, "\n")
cat("MAE: ", mae_train, "\n")


```

The ARDL predictions are not accurate on the training sample. 
Furthermore, the metric performances are not good. This can already imply that they won't perform well on the test sample. 

We can obbserve it by observing the result of this code : here, we perform the predictions on the test sample; then, we "convert" the returns to closing prices of the CAC 40 index thanks to the last closing price of the training sample. Then, we plot it. 
```{r}
# Prepare the dataframe for the test set
new_dataframe_test <- new_dataframe[new_dataframe$Date >= "2023-01-01", ]

# Add necessary lags for Return and CPIH_YTYPCT
new_dataframe_test <- new_dataframe_test %>%
  mutate(
    L1_Return = lag(Return, 1, default = tail(new_dataframe_train$Return, 1)),
    L2_Return = lag(Return, 2, default = tail(new_dataframe_train$Return, 2)[1]),
    L3_Return = lag(Return, 3, default = tail(new_dataframe_train$Return, 3)[1]),
    L1_CPIH_YTYPCT = lag(CPIH_YTYPCT, 1, default = tail(new_dataframe_train$CPIH_YTYPCT, 1)),
    L2_CPIH_YTYPCT = lag(CPIH_YTYPCT, 2, default = tail(new_dataframe_train$CPIH_YTYPCT, 2)[1]),
    L3_CPIH_YTYPCT = lag(CPIH_YTYPCT, 3, default = tail(new_dataframe_train$CPIH_YTYPCT, 3)[1])
  )

# Rename columns to match the ARDL model syntax
new_dataframe_test <- new_dataframe_test %>%
  rename(
    `L(Return, 1)` = L1_Return,
    `L(Return, 2)` = L2_Return,
    `L(Return, 3)` = L3_Return,
    `L(CPIH_YTYPCT, 1)` = L1_CPIH_YTYPCT,
    `L(CPIH_YTYPCT, 2)` = L2_CPIH_YTYPCT,
    `L(CPIH_YTYPCT, 3)` = L3_CPIH_YTYPCT
  )

# Make predictions using the ARDL model converted to a linear model
predictions_test <- predict(ardl_new_lm, newdata = new_dataframe_test)

# Add the predictions to the test dataframe
new_dataframe_test$Predicted <- predictions_test

# Calculate the errors
errors_test <- new_dataframe_test$Return - new_dataframe_test$Predicted

# Calculate performance metrics
rmse_test <- sqrt(mean(errors_test^2, na.rm = TRUE))
mae_test <- mean(abs(errors_test), na.rm = TRUE)

cat("Performance on the test set:\n")
cat("RMSE: ", rmse_test, "\n")
cat("MAE: ", mae_test, "\n")

# Plot predictions vs observed values
library(ggplot2)
ggplot(new_dataframe_test, aes(x = Date)) +
  geom_line(aes(y = Return), color = 'black', linewidth = 1, linetype = "solid") +  # Observed
  geom_line(aes(y = Predicted), color = 'blue', linewidth = 1, linetype = "dashed") +  # Predicted
  labs(
    title = "ARDL predictions on the test sample",
    x = "Date",
    y = "Return"
  ) +
  theme_minimal()

# Last known closing price (from the end of the training set)
last_train_price <- 6473.76

# Calculate observed and predicted prices
new_dataframe_test <- new_dataframe_test %>%
  mutate(
    Observed_Price = last_train_price * cumprod(exp(Return)),  # Observed prices
    Predicted_Price = last_train_price * cumprod(exp(Predicted))  # Predicted prices
  )

# Plot observed vs predicted prices
ggplot(new_dataframe_test, aes(x = Date)) +
  geom_line(aes(y = Observed_Price), color = 'black', linewidth = 1, linetype = "solid") +  # Observed prices
  geom_line(aes(y = Predicted_Price), color = 'blue', linewidth = 1, linetype = "dashed") +  # Predicted prices
  labs(
    title = "Observed vs Predicted Closing Prices",
    x = "Date",
    y = "Closing Price"
  ) +
  theme_minimal()

# Calculate errors between observed and predicted prices
errors_prices <- new_dataframe_test$Observed_Price - new_dataframe_test$Predicted_Price

# Calculate RMSE and MAE for the prices
rmse_prices <- sqrt(mean(errors_prices^2, na.rm = TRUE))  # RMSE
mae_prices <- mean(abs(errors_prices), na.rm = TRUE)      # MAE

# Calculate R-squared (coefficient of determination)
ss_total <- sum((new_dataframe_test$Observed_Price - mean(new_dataframe_test$Observed_Price, na.rm = TRUE))^2, na.rm = TRUE)  # Total sum of squares
ss_residual <- sum(errors_prices^2, na.rm = TRUE)  # Residual sum of squares
r_squared <- 1 - (ss_residual / ss_total)  # R-squared

# Display results
cat("Performance on closing prices:\n")
cat("RMSE (Price): ", rmse_prices, "\n")
cat("MAE (Price): ", mae_prices, "\n")
cat("R-squared (Price): ", r_squared, "\n")


````

We observe a decreasing trend, which is really not good because the time serie of the closing prices has a linear increasing trend. 
